---
title: "Applied Data Analysis"
subtitle: "Analyzing word and document frequency: tf-idf"
author: "Tomáš Oleš"
institute: "Department of Economic Policy"
date: "2026"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: atom-one-light
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
#See slide on xaringanthemer
library(xaringanthemer)
# style_duo_accent(
# primary_color = "#d19f2a",     # KHP color
# secondary_color = "#e2c47c",   # KHP light
style_mono_accent(
base_color = "#002147",
header_font_google = google_font("Josefin Sans"),
text_font_google = google_font("Montserrat", "300", "300i"),
code_font_google = google_font("Fira Mono")
)

```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
# Remember to compile
#xaringan::inf_mr(cast_from = "..")
#       slideNumberFormat: ""  
library(tidyverse)
if (!require("emo")) devtools::install_github("")
library(emo)
library(DiagrammeR)
library(Tmisc)
library(ggplot2)
library(gridExtra)
if (!require("dsbox")) devtools::install_github("rstudio-education/dsbox")
library(dsbox)
library(palmerpenguins)
library(tibble)
# Show only 5 rows of a tibble by default
options(tibble.print_max = 3, tibble.print_min = 3)
```

```{r  echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(fig.retina = 5)
```

# Motivation (i.)

A central question in **text mining** and **NLP** is:

> How can we quantify *what a document is about*?

---
---

class: middle

# Analyzing word and document frequency: tf-idf

---

## Word importance

- One idea: look at which words appear in a document  
- **Term frequency (tf):** how often a word occurs  
- But: some words appear in every document  
  - “the”, “is”, “of”, etc.  
- These are not informative!

---

## Beyond stop words

We can remove common words (stop words),  
but this is a crude approach.

> We need a dynamic way to **down-weight ubiquitous words**.

---

# Inverse document frequency (idf)

Another idea: give less weight to words  
that appear in many documents.

$$idf(\text{term}) = \ln{\left(\frac{n_{\text{documents}}}{n_{\text{documents containing term}}}\right)}$$

---

## Combining tf and idf

The **tf–idf** score = term frequency × inverse document frequency

- High tf–idf → frequent in this document, rare elsewhere  
- Low tf–idf → common word shared across documents  

---

## Interpretation

> tf–idf measures how important a word is to a document  
> within a collection of documents (the *corpus*).

It’s a **heuristic**—useful in search engines, topic modeling, and NLP.

---

# Example: Jane Austen’s novels

We’ll start with the published novels of **Jane Austen**  
and examine term frequency, then tf–idf.

---

## Data setup

```{r}
library(dplyr)
library(janeaustenr)
library(tidytext)

book_words <- austen_books() %>%
  unnest_tokens(word, text) %>%
  count(book, word, sort = TRUE)

total_words <- book_words %>% 
  group_by(book) %>% 
  summarize(total = sum(n))

book_words <- left_join(book_words, total_words)

book_words
```
---
### Term frequency distribution

```{r warning=FALSE, out.width='38%',fig.align="center"}
library(ggplot2)

ggplot(book_words, aes(n/total, fill = book)) +
  geom_histogram(show.legend = FALSE) +
  xlim(NA, 0.0009) +
  facet_wrap(~book, ncol = 2, scales = "free_y")
```

---
### Zipf’s Law

 > The frequency of a word is inversely proportional to its rank.

 This relationship is common in natural language corpora.
```{r warning=FALSE, out.width='38%',fig.align="center"}
freq_by_rank <- book_words %>%
  group_by(book) %>%
  mutate(rank = row_number(),
         term_frequency = n/total)
freq_by_rank
```
---
### Zipf's law

```{r warning=FALSE, out.width='50%',fig.align="center"}
freq_by_rank %>% 
  ggplot(aes(rank, term_frequency, color = book)) + 
  geom_line(linewidth = 1.1, alpha = 0.8, show.legend = FALSE) + 
  scale_x_log10() +
  scale_y_log10()
```



---
### Term frequency distribution

```{r}
rank_subset <- freq_by_rank %>%
  filter(rank < 500, rank > 10)

lm(log10(term_frequency) ~ log10(rank), data = rank_subset)
```


---
### Visualizing with the fitted line

```{r warning=FALSE, out.width='40%',fig.align="center"}
freq_by_rank %>% 
  ggplot(aes(rank, term_frequency, color = book)) + 
  geom_abline(intercept = -0.62, slope = -1.1, 
              color = "gray50", linetype = 2) +
  geom_line(linewidth = 1.1, alpha = 0.8, show.legend = FALSE) + 
  scale_x_log10() +
  scale_y_log10()
```

---

### Computing the tf-idf

Use `bind_tf_idf()` from `tidytext`.

```{r warning=FALSE, out.width='40%',fig.align="center"}
book_tf_idf <- book_words %>%
  bind_tf_idf(word, book, n)

# Inspect the tf-idf
book_tf_idf %>%
  arrange(desc(tf_idf))
```

---

```{r warning=FALSE, out.width='50%',fig.align="center"}
book_tf_idf %>%
  group_by(book) %>%
  slice_max(tf_idf, n = 15) %>%
  ungroup() %>%
  ggplot(aes(tf_idf, fct_reorder(word, tf_idf), fill = book)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~book, ncol = 2, scales = "free") +
  labs(x = "tf-idf", y = NULL)
```
---

### Example: classic physics texts 


Let’s try the same on scientific works:

 + Galileo: Discourse on Floating Bodies

 + Huygens: Treatise on Light

 + Tesla: Experiments with Alternate Currents

 + Einstein: Relativity: The Special and General Theory

```{r warning=FALSE, out.width='40%',fig.align="center"}
library(gutenbergr)

physics <- gutenberg_download(
  c(37729, 14725, 13476, 30155),
  meta_fields = "author"
)

physics_words <- physics %>%
  unnest_tokens(word, text) %>%
  count(author, word, sort = TRUE)
```

---

```{r warning=FALSE, out.width='40%',fig.align="center"}
physics_words
```

```{r}
plot_physics <- physics_words %>%
  bind_tf_idf(word, author, n) %>%
  mutate(author = factor(author, levels = c("Galilei, Galileo",
                                            "Huygens, Christiaan", 
                                            "Tesla, Nikola",
                                            "Einstein, Albert")))
```

---

```{r warning=FALSE, out.width='40%',fig.align="center"}
plot_physics %>% 
  group_by(author) %>% 
  slice_max(tf_idf, n = 15) %>% 
  ungroup() %>%
  mutate(word = reorder(word, tf_idf)) %>%
  ggplot(aes(tf_idf, word, fill = author)) +
  geom_col(show.legend = FALSE) +
  labs(x = "tf-idf", y = NULL) +
  facet_wrap(~author, ncol = 2, scales = "free")

```
---
--- 

```{r warning=FALSE, out.width='40%',fig.align="center"}
library(stringr)

physics %>% 
  filter(str_detect(text, "_k_")) %>% 
  select(text)
```

---
--- 
### Custom stop words removal
```{r warning=FALSE, out.width='40%',fig.align="center"}
mystopwords <- tibble(word = c("eq", "co", "rc", "ac", "ak", "bn", 
                                   "fig", "file", "cg", "cb", "cm",
                               "ab", "_k", "_k_", "_x"))

physics_words <- anti_join(physics_words, mystopwords, 
                           by = "word")

```

---
--- 

```{r warning=FALSE, out.width='30%',fig.align="center"}
plot_physics <- physics_words %>%
  bind_tf_idf(word, author, n) %>%
  mutate(word = str_remove_all(word, "_")) %>%
  group_by(author) %>% 
  slice_max(tf_idf, n = 15) %>%
  ungroup() %>%
  mutate(word = fct_reorder(word, tf_idf)) %>%
  mutate(author = factor(author, levels = c("Galilei, Galileo",
                                            "Huygens, Christiaan",
                                            "Tesla, Nikola",
                                            "Einstein, Albert")))

ggplot(plot_physics, aes(tf_idf, word, fill = author)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~author, ncol = 2, scales = "free") +
  labs(x = "tf-idf", y = NULL)
```
---
### Usecase in economics

- **Applications:**
  - **Labor economics:** Identify skill content in job vacancies or resumes (e.g., Deming & Kahn, 2018; Modestino et al., 2020).  
    → TF–IDF weights reflect demand intensity for each skill.
  - **Innovation and productivity:** Measure technological similarity between patents and firms’ text disclosures.  
    → Used to link innovation clusters and productivity spillovers.
  - **Political economy & finance:** Extract themes from speeches, central bank statements, or reports.  
    → TF–IDF helps detect attention shifts or sentiment changes over time.

- **Advantages:**
  - Simple, interpretable weighting scheme  
  - Reduces dominance of frequent but uninformative words  
  - Useful as input for clustering, topic modeling, or regression analysis
  
> After learning some regression techniques based on TF-IDF, we will classify emails as spam or ham (week 6). This can easily be extended to automation and non-automation innovations, as pioneered by Mann & Putman (2023).
---

# Sources

- [Deming, D. J., & Kahn, L. B. (2018). *Skill Requirements across Firms and Labor Markets: Evidence from Job Postings.* The Journal of Labor Economics.](https://doi.org/10.1086/694106)

- [Modestino, A. S., Shoag, D., & Ballance, J. (2020). *Downskilling: Changes in Employer Skill Requirements over the Business Cycle.* Labour Economics, 65, 101821.](https://doi.org/10.1016/j.labeco.2020.101821)

- [Mann, K., & Putman, L. (2023). *Benign Effects of Automation.* Working Paper, CBS.](https://research-api.cbs.dk/ws/portalfiles/portal/71105288/katja_mann_et_al_benign_effects_of_automation_acceptedversion.pdf)

- [Silge, J., & Robinson, D. (2017). *Text Mining with R: A Tidy Approach.* O'Reilly.](https://www.tidytextmining.com/)

---
