---
title: "Applied Data Analysis"
subtitle: "Clustering"
author: "Martin Vaško"
institute: "Department of Economic Policy"
date: "2026"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: atom-one-light
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(xaringanthemer)
style_mono_accent(
base_color = "#002147",
header_font_google = google_font("Josefin Sans"),
text_font_google = google_font("Montserrat", "300", "300i"),
code_font_google = google_font("Fira Mono")
)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(fig.retina = 5, message = FALSE, warning = FALSE)
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(tidyverse)
library(cluster)
library(factoextra)
options(dplyr.print_min = 8, dplyr.print_max = 8)
set.seed(2026)
```

---

## Lecture overview

1. **Unsupervised learning**: what changes when there is no $y$
2. **Similarity and distance**: how we measure "close"
3. **K-means clustering**: algorithm, objective, pitfalls
4. **Choosing the number of clusters**: elbow, silhouette, gap statistic
5. **Hierarchical clustering**: dendrograms and linkage
6. **Principal component analysis**: dimension reduction before clustering
7. **Application**: grouping economies by structural indicators

---

class: middle

# Unsupervised learning

---

## Supervised vs. unsupervised

So far every method had a **target variable**.

| | Supervised | Unsupervised |
|---|---|---|
| Data | $(X, y)$ | $X$ only |
| Goal | predict $y$ | find structure in $X$ |
| Success | test error, accuracy | no single ground truth |
| Examples | regression, logistic, KNN | **clustering**, PCA |

--

In unsupervised learning there is **no answer key**. We cannot compute a test error,
so the analysis is more *exploratory* — and the results need more judgement.

---

## Why economists cluster

- **Regional economics**: group regions into development types before modelling convergence
- **Labour economics**: segment workers by skill profiles from survey data
- **Firm-level analysis**: identify peer groups of firms with similar cost structures
- **Macro**: classify countries by institutional or structural similarity
- **Finance**: group assets by co-movement to build diversified portfolios

> Clustering is often a **pre-processing step**: the cluster label becomes a variable
> in a later regression.

---

class: middle

# Similarity and distance

---

## The notion of similarity

Clustering needs a formal answer to: *how far apart are two observations?*

For observations $x_i, x_j \in \mathbb{R}^p$:

**Euclidean distance**
$$d(x_i, x_j) = \sqrt{\sum_{k=1}^{p} (x_{ik} - x_{jk})^2}$$

**Manhattan distance**
$$d(x_i, x_j) = \sum_{k=1}^{p} |x_{ik} - x_{jk}|$$

**Correlation-based distance** treats two observations as similar if their *profiles*
across variables move together, regardless of magnitude.

---

## Scaling is not optional

Distance is dominated by whichever variable has the largest units.

```{r}
data("USArrests")
round(sapply(USArrests, sd), 2)
```

`Assault` has a standard deviation roughly **80 times** that of `Murder`. Without
scaling, the clustering would be an analysis of assault counts and nothing else.

```{r}
arrests <- scale(USArrests)
round(sapply(as.data.frame(arrests), sd), 2)
```

> Rule of thumb: **standardise** unless the variables are already in comparable units.

---

## Looking at the distance matrix

```{r out.width='52%', fig.align="center", fig.width=7, fig.height=6}
fviz_dist(dist(arrests), show_labels = FALSE) +
  labs(title = "Pairwise Euclidean distance, scaled US arrest data")
```

---

class: middle

# K-means clustering

---

## The K-means objective

Partition observations into $K$ non-overlapping clusters $C_1, \dots, C_K$ that
minimise the **total within-cluster variation**:

$$\min_{C_1,\dots,C_K} \sum_{k=1}^{K} W(C_k)$$

Using squared Euclidean distance, the within-cluster variation is

$$W(C_k) = \frac{1}{|C_k|} \sum_{i, i' \in C_k} \sum_{j=1}^{p} (x_{ij} - x_{i'j})^2
 = 2 \sum_{i \in C_k} \sum_{j=1}^p (x_{ij} - \bar{x}_{kj})^2$$

where $\bar{x}_{kj}$ is the mean of variable $j$ in cluster $k$ — the **centroid**.

> Searching all partitions is infeasible: there are roughly $K^n$ of them.
> K-means uses a greedy algorithm instead.

---

## The algorithm

1. Randomly assign each observation to one of $K$ clusters.
2. Repeat until assignments stop changing:
   - **Update step**: compute the centroid of each cluster.
   - **Assign step**: assign each observation to the nearest centroid.

--

Each step can only **decrease** the objective, so the algorithm always converges.

.pull-left[
**Guaranteed**: convergence to *a* local optimum
]
.pull-right[
**Not guaranteed**: the *global* optimum
]

> Because the starting assignment is random, different runs can give different answers.
> Always use `nstart` to run the algorithm many times and keep the best result.

---

## K-means in R

```{r}
km4 <- kmeans(arrests, centers = 4, nstart = 25)
km4$size
round(km4$centers, 2)
```

---

## Visualising the clusters

```{r out.width='55%', fig.align="center", fig.width=8, fig.height=5.5}
fviz_cluster(km4, data = arrests, repel = TRUE, labelsize = 8) +
  labs(title = "K-means, K = 4")
```

---

## Why `nstart` matters

```{r}
set.seed(1)
bad  <- kmeans(arrests, centers = 4, nstart = 1)
good <- kmeans(arrests, centers = 4, nstart = 50)

c(nstart_1 = bad$tot.withinss, nstart_50 = good$tot.withinss)
```

A single random start can land in a noticeably worse local optimum.

> Practical advice: `nstart = 25` or higher, and `set.seed()` for reproducibility.

---

class: middle

# How many clusters?

---

## The elbow method

Plot total within-cluster sum of squares against $K$ and look for the kink.

```{r out.width='50%', fig.align="center", fig.width=7, fig.height=4.5}
fviz_nbclust(arrests, kmeans, method = "wss") +
  labs(subtitle = "Elbow method")
```

---

## The silhouette

For observation $i$, let $a(i)$ be its mean distance to other points in its own
cluster and $b(i)$ the mean distance to the nearest *other* cluster:

$$s(i) = \frac{b(i) - a(i)}{\max\{a(i), b(i)\}} \in [-1, 1]$$

- $s(i)$ near $1$: well clustered
- $s(i)$ near $0$: on the boundary between two clusters
- $s(i) < 0$: probably in the wrong cluster

```{r out.width='45%', fig.align="center", fig.width=7, fig.height=3.6}
fviz_nbclust(arrests, kmeans, method = "silhouette") +
  labs(subtitle = "Average silhouette width")
```

---

## The gap statistic

Compares the observed within-cluster dispersion to the dispersion expected under a
**null reference distribution** with no clustering at all.

```{r out.width='50%', fig.align="center", fig.width=7, fig.height=4.2}
gap <- clusGap(arrests, FUN = kmeans, nstart = 25, K.max = 8, B = 50)
fviz_gap_stat(gap)
```

---

## Reading the three criteria together

The three methods frequently **disagree** — and that is informative, not a failure.

- The elbow is often ambiguous on real data
- The silhouette favours few, well-separated clusters
- The gap statistic can favour $K = 1$ when structure is weak

> Choose $K$ with the **research question** in mind. If you need four policy groups,
> a defensible $K = 4$ beats a statistically marginal $K = 2$.

---

class: middle

# Hierarchical clustering

---

## No need to fix K in advance

**Agglomerative** (bottom-up) hierarchical clustering:

1. Start with each observation as its own cluster.
2. Repeatedly merge the two **closest** clusters.
3. Stop when one cluster remains.

The result is a **dendrogram**: a tree that records the whole merge history.
Cutting the tree at a given height yields a clustering — one tree gives you
every $K$ at once.

---

## Linkage: distance between *clusters*

Distance between two observations is clear. Between two **groups** it is a choice.

| Linkage | Definition | Behaviour |
|---|---|---|
| **Complete** | largest pairwise distance | compact, balanced clusters |
| **Average** | mean pairwise distance | compact, balanced clusters |
| **Single** | smallest pairwise distance | "chaining", trailing clusters |
| **Centroid** | distance between centroids | can produce inversions |

> Complete and average linkage are the usual defaults; single linkage tends to
> produce long straggly clusters.

---

## Building the dendrogram

```{r out.width='58%', fig.align="center", fig.width=10, fig.height=5}
hc <- hclust(dist(arrests), method = "complete")

fviz_dend(hc, k = 4, cex = 0.55, rect = TRUE,
          main = "Complete linkage, scaled US arrest data")
```

---

## Cutting the tree

```{r}
grp <- cutree(hc, k = 4)
table(grp)
split(rownames(USArrests), grp)[[1]]
```

```{r}
# do the two methods agree?
table(hierarchical = grp, kmeans = km4$cluster)
```

---

## Comparing linkages

```{r out.width='70%', fig.align="center", fig.width=12, fig.height=4}
par(mfrow = c(1, 3), cex = 0.45)
plot(hclust(dist(arrests), "complete"), main = "Complete", xlab = "", sub = "")
plot(hclust(dist(arrests), "average"),  main = "Average",  xlab = "", sub = "")
plot(hclust(dist(arrests), "single"),   main = "Single",   xlab = "", sub = "")
```

Note the chaining in the single-linkage tree.

---

class: middle

# Principal component analysis

---

## Why reduce dimension first

With $p$ large:

- distances become less informative (everything is far from everything)
- we cannot plot the clusters
- correlated variables are effectively **counted twice** in the distance

**PCA** finds a small number of directions that capture most of the variance.

The first principal component is the normalised linear combination

$$Z_1 = \phi_{11} X_1 + \phi_{21} X_2 + \dots + \phi_{p1} X_p$$

with $\sum_j \phi_{j1}^2 = 1$ that has the **largest possible variance**. $Z_2$ is the
next such direction, **orthogonal** to $Z_1$, and so on.

---

## PCA in R

```{r}
pca <- prcomp(USArrests, scale. = TRUE)
round(pca$rotation, 3)
```

PC1 loads on all four crime variables with similar weight — an overall **crime rate**.
PC2 is dominated by `UrbanPop` — **urbanisation**.

---

## How much variance is explained?

```{r out.width='48%', fig.align="center", fig.width=7, fig.height=4}
summary(pca)$importance[, 1:4] |> round(3)
fviz_eig(pca, addlabels = TRUE, barfill = "#002147", barcolor = "#002147")
```

---

## The biplot

```{r out.width='55%', fig.align="center", fig.width=8, fig.height=5.5}
fviz_pca_biplot(pca, repel = TRUE, labelsize = 3,
                col.var = "#b22222", col.ind = "#002147")
```

---

## Clustering in PCA space

A common workflow: **PCA first, then cluster the leading components.**

```{r out.width='50%', fig.align="center", fig.width=8, fig.height=4.6}
scores  <- pca$x[, 1:2]
km_pca  <- kmeans(scores, centers = 4, nstart = 25)

fviz_cluster(km_pca, data = scores, repel = TRUE, labelsize = 8) +
  labs(title = "K-means on the first two principal components")
```

---

## Interpreting the groups

```{r}
USArrests |>
  as_tibble(rownames = "state") |>
  mutate(cluster = factor(km4$cluster)) |>
  group_by(cluster) |>
  summarise(n = n(), across(Murder:Rape, ~ round(mean(.x), 1)))
```

> Clusters are only useful once you can **name** them. Profile each cluster on the
> original variables, not the scaled ones.

---

class: middle

# Practical caveats

---

## What to watch out for

* **Clustering always returns clusters.** The algorithm will happily partition pure
  noise. Validate before you interpret.
* **Scaling drives the result.** Report what you scaled and why.
* **K-means assumes roughly spherical, equally sized clusters.** It struggles with
  elongated or nested shapes.
* **Outliers distort centroids.** Consider `pam()` (k-medoids), which is more robust.
* **The choice of linkage and distance is a modelling assumption**, not a technicality.
* **Results are not a test.** There is no p-value here — clustering generates
  hypotheses, it does not confirm them.

---

## A robustness check you should always run

```{r out.width='45%', fig.align="center", fig.width=7, fig.height=3.6}
# k-medoids: medoids are actual observations, so outliers matter less
pam4 <- pam(arrests, k = 4)
table(kmeans = km4$cluster, pam = pam4$clustering)
```

If the partition collapses under a small change of method, it was never stable.

---

# Sources

- [James, G., Witten, D., Hastie, T., Tibshirani, R. (2021). *An Introduction to Statistical Learning*, Ch. 12. Springer.](https://www.statlearning.com/)

- [Kassambara, A., Mundt, F. *factoextra: Extract and Visualize the Results of Multivariate Data Analyses.*](https://cran.r-project.org/package=factoextra)

- [Hastie, T., Tibshirani, R., Friedman, J. (2009). *The Elements of Statistical Learning*, Ch. 14.](https://hastie.su.domains/ElemStatLearn/)

- Maechler, M. et al. *cluster: Cluster Analysis Basics and Extensions.*

---
